---
title: "Temperate-Tropical Split"
author: "Connor French"
date: "7/23/2020"
output: html_document
---

I'm splitting the globe into a temperate-tropical divide based on: 
> White, A.E., Dey, K.K., Mohan, D., Stephens, M., Price, T.D., 2019. Regional influences on community structure across the tropical-temperate divide. Nat. Commun. 10, 2646. https://doi.org/10.1038/s41467-019-10253-6.  

They observe a sharp overturn in community composition along a "freeze line", where the minimum temperature of the coldest month is below zero degrees. This corresponds roughly with temperate and tropical biomes. I am using their delineation and further refining the distinction by considering any "tropical" areas defined by the freeze line that overlap with temperate biomes as temperate. The code for creating a binary raster from the minimum temperature of the coldest month divide is in the "step-0_aggregation.ipynb" script. 


## Setup
```{r, setup}
library(tidyverse)
library(stars)
library(sf)
library(here)
```


Read in the file and get a rough plot.
```{r}
binary_raster <- read_stars(here("data", "climate_poly", "min_temp_binary.tif"))

plot(binary_raster)
```

Convert to a polygon.
```{r}
binary_poly <- st_as_sf(binary_raster)

binary_poly
```

Group the polygons by the binary classification and simplify to two polygons
```{r}
binary_union <- binary_poly %>% 
  mutate(min_temp = if_else(min_temp_binary.tif > 0, "tropical", "temperate")) %>% 
  group_by(min_temp) %>% 
  summarize(is_tropical = mode(min_temp_binary.tif))


```


```{r}
plot(binary_union)
```


